Full-day, in home validation of infant body position measurements from inertial sensors

### 2023-03-14 manuscript.Rmd
edited intro

From moment to moment, infants' movements facilitate and constrain how they can interact with their surroundings. Changes in *body position*---whether infants are supine on their backs, prone on their bellies, sitting, upright, or held by a  caregiver---have immediate consequences for vision, object exploration, and social interaction. When sitting and upright, infants have a better view of faces and distant objects compared to their view while in a prone position [@Freeplay; @LuoFranchak2020; @CWW]. Infants struggle to manipulate objects while supine and prone, but sitting affords object exploration [@SoskaAdolph2014]. Upright walking changes how infants and caregivers interact compared with crawling in a prone position; while walking infants move farther away, share toys in different ways, and hear different language from caregivers [@ChenSchneider2022; @Karasik2011; @WestIverson2021; @Karasik2014]. As infants grow older and acquire new abilities, such as independent sitting and walking, the spend more time sitting and upright and less time held, supine, and prone [@Survey; @Thurman2017; @Freeplay; @AdolphCTL2014]. Thus, characterizing individual differences in the day-to-day accumulation of body position experiences informs developmental theory by revealing differential opportunities for learning [@CurrentOpinion].
<!-- Other west cites; cite new EMA; KretchKoziol2022 -->

In this paper, we present an inertial sensing method to classify infant body position from moment-to-moment across an entire day, and validate its accuracy using over 100 hours of video recorded across 35 in-home data collection sessions. Our method takes inspiration from a more mature technology: Long-form audio recordings of infants' language experiences using wearable audio recorders. We begin by describing the impact of long-form audio recordings on research in developmental psychology, and identify the key features that should be replicated in long-form recordings of motor behavior. Next, we review the current state-of-the-art in measuring infant motor behavior---video and survey data---and their limitations in capturing real-time, full-day behavior. Finally, we discuss the advantages of using inertial sensing to classify motor behavior. Despite promising past results in brief, supervised sessions [@FranchakScott2021; @AiraksinenRasanen2020; @AiraksinenGallen2022], the current investigation takes a needed step forward by testing accuracy over long, unsupervised recordings. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion [@AiraksinenRasanen2020] and infant-caregiver contact [@YaoPlotz2019]. 
<!-- Accompanying our paper, we have created a database [CITE] containing synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.  -->

@@

The LENA&reg; recorder is a commercial device that is worn by infants in a custom shirt pocket that has sufficient battery life and storage to record for an entire day. Closed-source LENA&reg; algorithms analyze the audio recordings to provide automatic counts of useful metrics, such as the number of words spoken by adults in the vicinity of the participant. Other long-form audio methods rely on custom-built recorders [@WassPhillips2022], apply alternative classification algorithms to LENA&reg; data [@RasanenSeshadri2020; @MichelettiYao2022], or manually transcribe audio recorded by LENA&reg; devices to improve accuracy or identify behaviors beyond the built-in categories [@Bergelson2019Corpus; @MendozaFausey2021].

Long-form audio recordings have had a transformational impact on language development research by allowing researchers to characterize opportunities for learning in daily life. Measuring the amount of speech heard by infants in the home [@Weisleder2013] or in a daycare setting [@PerryPrince2018] revealed individual differences input that predict later vocabulary. Full-day language recording synchronized with other data sources allows researchers to identify how auditory input and vocal production interact with other processes. Beyond individual differences in aggregated data, long-form recordings can be used to determine the temporal schedule of experiences. For example, infants' daily experiences hearing music are clustered in time, with "bursty" episodes of hearing music separated by relatively long periods during which music is absent [@MendozaFausey2022]. Synchronizing audio recordings with other data sources extends researchers' ability to characterize daily experiences. Linking LENA&reg; speech measurements to repeated, time-stamped text-message surveys about infant device placement revealed that infants heard less caregiver speech during moments that they were restrained in devices such as swings, exersaucers, and car seats [@MalachowskiSalo2023]. A custom-built wearable ECG and audio recorder allowed @WassPhillips2022 to discover that infant arousal increases the likelihood of infant vocalization across the day.

We identified five key features of long-form audio methods that should be replicated in analogous studies of motor behavior. First, wearable audio recorders are *mobile*. Measurement is not limited to a particular room because the recording device travels with the participant. Data are recorded to onboard device memory, so participants do not need to be in range of a receiver. Second, wearable audio recording is *unobtrusive*. Participants' reactivity to observation, such as from a video camera, may influence behavior. For example, caregivers spoke more frequently to infants during a video-recorded portion of a home recording compared with audio-only segments captured by a LENA&reg; device [@Bergelson2019Input]. Third and fourth, recordings capture *real-time data* over a *full day*. The ability to record real-time data is vital for making inferences about processes that happen on the timescale of minutes or even seconds within in an individual as opposed to comparisons of aggregated data between infants. Synchronizing real-time data to other data streams helps to reveal sources of variability within an individual [e.g., @WassPhillips2022]. Full-day recordings are essential for capturing experiences across the heterogeniety of daily routines that moderate behavior (e.g., play, feeding, errands) [@KadookaCaufield2021; @Tamis-LeMondaCustode2018]. "Burstiness" of behavior means that long recordings are needed to capture clusters of events amid long periods in which they may be absent [@BarbaroFausey2022; @WarlaumontSobowale2021]. Fifth, *automatic classification* means that the approach can scale to analyze large numbers of participants over long recordings without the bottleneck of manual annotation/transcription. Automatic classification can only replace human annotation if it is sufficiently accurate and unbiased. An independent assessment of the LENA&reg; algorithms found mixed results about the accuracy of different outcomes. For example, correlations between human transcribed counts of adult words and child vocalizations against LENA&reg;'s automatic counts were strong, *r* = .698 and *r* = .649, respectively [@CristiaLavechin2020]. However, poor agreement was found for other metrics, such as the number of "conversational turns" between the child and communicative partners and the identification of male speakers. 
<!-- EVENTUALLY CITE EMA HERE -->

Thus, for some use cases (and for some metrics), long form audio recordings provide a mobile, unobtrusive way to automatically score real-time data over a full day. In the remainder of the paper, we turn to the question of how to replicate these qualities in long form recordings of motor behavior.

@@

Video and survey methods are the current state-of-the-art in assessing motor behavior. Although, each method has  advantages and disadvantages for characterizing infants' everyday motor experiences that complements the other, neither method on its own can provide comparable data to the long form audio recordings reviewed in the previous section.

Video observation is the most common way of measuring infant motor behavior in home recordings. Most often, an experimenter with a handheld camera follows infants from room to room to ensure that their movements are visible throughout the recording session [@Karasik2011; @ChenSchneider2022; @HerzbergFletcher2021]. The primary advantage of video recording is that it captures real-time behavior. Infants' body position, locomotion, and reaching comprise events that occur on the timescale of seconds, so standard video recording is adequate to score gross motor behavior. However, requiring an experimenter to operate a camera is obtrusive, whereas relying on a stationary camera means that infants will be absent from view as they move from place to place. Moreover, video observation cannot easily scale to long durations or large numbers of participants. Logistically, an experimenter cannot follow behind infants to record their behavior from morning to night (and were they to do so, they would likely alter infants' and caregivers' behavior). Typical video recording sessions last 45-120 minutes [@Karasik2011; @ChenSchneider2022; @HerzbergFletcher2021], short of capturing the variety of activities across the full daily routine. Even if full day videos were available, the lack of suitable automatic classification tools means that the human cost of annotation makes it difficult for video methods to scale to testing large numbers of participants. Our annotation of body position takes approximately 2-5 hours to complete for every hour of video (depending on how often infants switch positions), meaning that a full "waking day" of approximately 11 hours for a 12-month-old [@GallandTaylor2012] could take 22-55 hours of labor to full annotate.

@@

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home [@CliffReilly2009; @Barbaro2019; @BruijnsTruelove2020; @LoboHall2019]. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commercially-available and inexpensive IMUs have > 12 hour battery life with onboard storage to record *real-time*, *full-day* motion data at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Moreover, past work has successfully recorded the rate of leg kicks [e.g., @DengTrujillo-Priego2019] and activity intensity [@SchnellerBentsen2017] in infants and children across multiple days.

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurement validity is acceptable over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity [e.g., @HagerTilton2017; @ArmstrongCovington2019].

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in "tummy time", or they can crawl in myriad ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, in a "W" position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019]. Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [@Preece2009; @Arif2015], children [@NamPark2013; @RenDing2016; @StewartNarayanan2018], and infants [@YaoPlotz2019; @FranchakScott2021; @AiraksinenRasanen2020]. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest model, that can be later used to predict categories from IMU data that was not annotated. Crucially, this allows automatic classification to scale to full day recording by relying on a relatively smaller set of video annotation.

<!-- NEED TO FIT IN GREENSPAN posture from cut points, Hewitt Tummy time, Yao holding-->

@@

<!-- Could have a bit more detail in this last paragraph -->

>>>>

From moment to moment, infants' movements facilitate and constrain how they can interact with their surroundings. Changes in *body position*---whether infants are supine on their backs, prone on their bellies, sitting, upright, or held by a caregiver---have immediate consequences for vision, object exploration, and social interaction. When sitting and upright, infants have a better view of faces and distant objects compared to their view while in a prone position [@Freeplay; @LuoFranchak2020; @CWW]. Infants struggle to manipulate objects while supine and prone, but sitting affords object exploration [@SoskaAdolph2014]. Upright walking changes how infants and caregivers interact; while walking infants move farther away from caregivers and share toys in different ways compared to when crawling in a prone position [@ChenSchneider2022; @Karasik2011]. As infants grow older and acquire new abilities, such as independent sitting and walking, they spend more time sitting and upright and less time held, supine, and prone [@Survey; @Thurman2017; @Freeplay; @AdolphCTL2014]. Caregivers provide different learning input to infants depending on infants' body position [@KretchKoziol2022; @WestFletcher2022; @WestIverson2021; @Karasik2014]. Thus, characterizing individual differences in the day-to-day accumulation of body position experiences informs developmental theory by revealing differential opportunities for learning [@CurrentOpinion].

In this paper, we present an inertial sensing method to classify infant body position from moment-to-moment across an entire day, and validate its accuracy using over 100 hours of video recorded across 34 in-home data collection sessions. Our method takes inspiration from a more mature technology: Long-form audio recordings of infants' language experiences using wearable audio recorders. We begin by describing the impact of long-form audio recordings on research in developmental psychology, and identify the key features that should be replicated in long-form recordings of motor behavior. Next, we review the current state-of-the-art in measuring infant motor behavior---video and survey data---and their limitations in capturing real-time, full-day behavior. Finally, we discuss the advantages of using inertial sensing to classify motor behavior. Despite promising past results in brief, supervised sessions [@FranchakScott2021; @AiraksinenRasanen2020; @AiraksinenGallen2022], the current investigation takes a needed step forward by testing accuracy over long, unsupervised recordings. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion [@AiraksinenRasanen2020] and infant-caregiver contact [@YaoPlotz2019]. 
<!-- Accompanying our paper, we have created a database [CITE] containing synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.  -->

@@

The LENA&reg; recorder is a commercial device that is worn by infants in a custom shirt pocket; the recorder has sufficient battery life and storage to record for an entire day. Closed-source LENA&reg; algorithms analyze the audio recordings to provide automatic counts of useful metrics, such as the number of words spoken by adults in the vicinity of the infant. Other long-form audio methods rely on custom-built recorders [@WassPhillips2022], apply alternative classification algorithms to LENA&reg; data [@RasanenSeshadri2020; @MichelettiYao2022], or manually transcribe audio recorded by LENA&reg; devices to improve accuracy or identify behaviors beyond the built-in categories [@Bergelson2019Corpus; @MendozaFausey2021].

Long-form audio recordings have had a transformational impact on language development research by allowing researchers to characterize opportunities for learning in daily life. For example, measuring the amount of speech heard by infants in the home [@Weisleder2013] or in a daycare setting [@PerryPrince2018] revealed how individual differences in speech input predict later vocabulary. Full-day language recording synchronized with other data sources allows researchers to identify how auditory input and vocal production interact with other processes. Beyond individual differences in aggregated data, long-form recordings can be used to determine the temporal schedule of experiences. For example, infants' daily experiences hearing music are clustered in time, with "bursty" episodes of music presence separated by relatively long periods during which music is absent [@MendozaFausey2022]. Synchronizing audio recordings with other data sources extends researchers' ability to characterize daily experiences. Linking LENA&reg; speech measurements to repeated, time-stamped text-message surveys about infant device placement revealed that infants heard less caregiver speech during periods that they were restrained in devices such as swings, exersaucers, and car seats [@MalachowskiSalo2023]. A custom-built wearable ECG and audio recorder allowed @WassPhillips2022 to discover that infant arousal increases the likelihood of infant vocalization across the day.

We identified five key features of long-form audio methods that should be replicated in analogous studies of motor behavior. First, wearable audio recorders are *mobile*. Measurement is not limited to a particular room because the recording device travels with the participant. Data are recorded to onboard device memory, so participants do not need to be in range of a receiver. Second, wearable audio recording is *unobtrusive*. Participants' reactivity to observation, such as from a video camera, may influence behavior. For example, caregivers spoke more frequently to infants during a video-recorded portion of a home recording compared with audio-only segments captured by a LENA&reg; device [@Bergelson2019Input]. Third and fourth, recordings capture *real-time data* over a *full day*. The ability to record real-time data is vital for making inferences about processes that happen on the timescale of minutes or even seconds within in an individual as opposed to comparisons of aggregated data between infants. Synchronizing real-time data to other data streams helps to reveal sources of variability within an individual [e.g., @WassPhillips2022]. Full-day recordings are essential for capturing experiences across the heterogeneity of daily routines that moderate behavior (e.g., play, feeding, errands) [@KadookaCaufield2021; @Tamis-LeMondaCustode2018]. "Burstiness" of behavior means that long recordings are needed to capture clusters of events amid long periods in which they may be absent [@BarbaroFausey2022; @WarlaumontSobowale2021]. Fifth, *automatic classification* means that the approach can scale to analyze large numbers of participants over long recordings without the bottleneck of manual annotation/transcription. But automatic classification can only replace human annotation if it is sufficiently accurate and unbiased. An independent assessment of the LENA&reg; algorithms found mixed results about the accuracy of different outcomes. For example, correlations between human transcribed counts of adult words and child vocalizations against LENA&reg;'s automatic counts were strong, *r* = .698 and *r* = .649, respectively [@CristiaLavechin2020]. However, poor agreement was found for other metrics, such as the number of "conversational turns" between the child and communicative partners and the identification of male speakers. 
<!-- EVENTUALLY CITE NEW EMA HERE -->

Thus, for some use cases (and for some metrics), long form audio recordings provide a mobile, unobtrusive way to automatically score real-time data over a full day. In the remainder of the paper, we turn to the question of how to replicate those key features in long form recordings of motor behavior.

@@

Video and survey methods are the current state-of-the-art in assessing motor behavior. Although each method has complementary advantages and disadvantages for characterizing infants' everyday motor experiences, neither method on its own can provide comparable data to the long form audio recordings reviewed in the previous section.

Video observation is the most common way of measuring infant motor behavior in home recordings. Most often, an experimenter with a handheld camera follows infants from room to room to ensure that their movements are visible throughout the recording session [@Karasik2011; @ChenSchneider2022; @HerzbergFletcher2021]. The primary advantage of video recording is that it captures real-time behavior. Infants' body position, locomotion, and reaching comprise events that occur on the timescale of seconds, so standard video recording is adequate to score gross motor behavior. However, requiring an experimenter to operate a camera is obtrusive, whereas relying on a stationary camera means that infants will be absent from view as they move from place to place. Moreover, video observation cannot easily scale to long durations or large numbers of participants. Logistically, an experimenter cannot follow behind infants to record their behavior from morning to night (and were they to do so, they would likely alter infants' and caregivers' behavior). Typical video recording sessions last 45-120 minutes [@Karasik2011; @ChenSchneider2022; @HerzbergFletcher2021], far short of capturing the variety of activities across the full daily routine. Even if full day videos were available, the lack of suitable automatic classification tools means that the human cost of annotation makes it difficult for video methods to scale to testing large numbers of participants. Our annotation of body position takes approximately 2-5 hours to complete for every hour of video (depending on how often infants switch positions), meaning that a full "waking day" of approximately 11 hours for a 12-month-old [@GallandTaylor2012] could take 22-55 hours of labor to full annotate.

@@

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home [@CliffReilly2009; @Barbaro2019; @BruijnsTruelove2020; @LoboHall2019]. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commercially-available and inexpensive IMUs have > 12 hour battery life with onboard storage to record *real-time*, *full-day* motion data at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Indeed, past work has successfully recorded the rate of leg kicks [e.g., @DengTrujillo-Priego2019] and activity intensity [@SchnellerBentsen2017] in infants and children across multiple days.

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity [e.g., @HagerTilton2017; @ArmstrongCovington2019].

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in "tummy time", or they can crawl in myriad ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, in a "W" position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019]. Although @GreenspanCunha2021 were able to use cut-point definitions about infant trunk angle to categorize body positions, this method was not able to differentiate times that infants were held by caregivers off the ground. Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [@Preece2009; @Arif2015], children [@NamPark2013; @RenDing2016; @StewartNarayanan2018], and infants [@YaoPlotz2019; @FranchakScott2021; @AiraksinenRasanen2020]. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest model, that can be later used to predict categories from IMU data that was not annotated. Crucially, this allows automatic classification to scale to full day recording by relying on a relatively smaller set of video annotation.

@@
